---
title: "lab 17"
author: "iman syed 18596789"
format: html
---



##Section 1: Identify genetic variants of interest

There are a number of gene variants associated with childhood asthma, it is important that we learn about them so we can can understand the disease better. A study from Verlaan
et al. (2009) shows that 4 candidate SNPs demonstrate significant evidence for association.
You want to find out what they are by visiting OMIM (http://www.omim.org) and locating the
Verlaan et al. paper description.

Here we read the CSV file drived from Ensembl


```{r}
mxl <- read.csv("MXL.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```
Let us look at a separate population:

Read in the file for GBR


```{r}
gbr <- read.csv("GBR.csv")
head(gbr)
```


Find proporton of G|G

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)* 100, 2 )
```


The variant that is associated with childhood asthma is more frequent in GBR populations than the MKL population.


##Section 4 :Population scale analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes:

```{r}
expres <-read.table("rs8067378_ENSG00000172057.6.txt")
head(expres)

```

```{r}
nrow(expres)
```
A13: a) There are 462 samples


```{r}
table(expres$geno)
```
A13: b) median expression levels for each of these genotype  A/A 108   A/G    233  G/G 121

```{r}
library(ggplot2)
```




let's make a plot:

Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
ggplot(expres)+ aes(geno,exp,fill=geno)+
  geom_boxplot(notch= TRUE)
```


A14: We can see that this location the relative expression value of the A/A genotype is the highest whilst G/G has the lowest expression value. This suggests that this SNP does affect the expression of the ORMDL3.
 










