---
title: "Class 10 mini project"
author: "Iman Syed 18596789"
format: pdf


---


As it is nearly halloween and the way point in the quarter let's do a mini project to help us find the best candy!

Ours comes from the 538 website and is available as a CSV file:

```{r}

candy <- read.csv ("candy-data.csv", row.names= 1)

head(candy)
```
Q1. How many different candy types are in this dataset?


```{r}
library (tidyverse)
candy |>
  nrow()
```


Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

My favorite winpercent

```{r}
candy["Almond Joy",]$winpercent
```

```{r}

library(dplyr)

candy|>
  filter (rownames(candy)== "Almond Joy") |>
  select(winpercent)



```

##Quick overview of the dataset

```{r}

skimr::skim(candy)

```


Q3. What is your favorite candy in the dataset and what is it’s winpercent value?
Q4. What is the winpercent value for “Kit Kat”?
Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
Q7. What do you think a zero and one represent for the candy$chocolate column?



Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy)+
  aes(winpercent)+
  geom_histogram(bins=20)
```


Q9. Is the distribution of winpercent values symmetrical?

```{r}
ggplot(candy)+
  aes(winpercent)+
  geom_density()
```


Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```


```{r}
summary(candy$winpercent)
```


Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#1.Find all chocolates candy in the dataset
#2 Find their winpercent values
#3 Calculate the mean of these values
#4-6. Do the same for fruity candy
#7 Compare mean winpercent of the chocolate vs. fruity
#8 Pick the highest as the winner

choc.inds <- candy$chocolate==1
choc.wins <- candy[choc.inds,]$winpercent
choc.mean <- mean(choc.wins)
choc.mean

```
```{r}
mean(candy[candy$chocolate==1,]$winpercent)
```
```{r}
mean(candy[candy$fruity==1,]$winpercent)

fruit.ind <- candy$fruity==1
fruit.wim <- candy[fruit.ind,]$winpercent
fruit.mean <- mean(fruit.wim)
fruit.mean

```

```{r}
candy|>
  filter(chocolate==1) |>
  select(winpercent)

```



Q12. Is this difference statistically significant?

```{r}
t.test(choc.wins, fruit.wim)


```



13.

```{r}
candy|>
  arrange(winpercent) |>
  head(5)

```


```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,],5)
```



15. make a first barplot of candy ranking based on winpercent values:

```{r}
ggplot(candy)+
  aes(winpercent,rownames(candy))+
  geom_col()

```



16.
```{r}
ggplot(candy) +
  aes(x = winpercent, y = reorder(rownames(candy), winpercent)) +
  geom_col()

```

add some color base on

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] <- "chocolate"
my_cols[as.logical(candy$frutiy)] <- "purple"
my_cols[as.logical(candy$bar)] <- "blue"
my_cols

```



```{r}

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + 
  

```
Now, for the first time, using this plot we can answer questions like:
- Q17. What is the worst ranked chocolate candy?
Whoppers!

- Q18. What is the best ranked fruity candy?

Airheads!



##Winpercent and Pricepercent
A plot  with both variables/columns winpercent and pricepercent

```{r}
ggplot(candy)+
  aes(x=winpercent, y=pricepercent,
      label= rownames(candy)) +
  geom_point(col=my_cols)+
  geom_text()
```






```{r}

library(ggrepel)

ggplot(candy)+
  aes(x=winpercent, y=pricepercent,
      label= rownames(candy)) +
  geom_point(col=my_cols)+
  geom_text_repel()

```

##Exploring correlation structure




```{r}
library(corrplot)
cij <- cor(candy)
cij
```

##Principal Component Analysis

The function to use is called 'prcom()' with an optional "scale" arguement

```{r}
pca<-prcomp(candy, scale= TRUE)

summary(pca)
```

```{r}
ggplot(pca$x)+
  aes(PC1, PC2, label= rownames(pca$x))+
  geom_point(col= my_cols)+
  geom_text(col=my_cols)

```

We should also examine the variable "loadings" or contributions of the original variables to new PCs

```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation))+
  geom_col()
```

##5 Exploring the correlation structure
Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.


```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```






```







